---
title: "Coaching Networks and Career Pathways Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(lubridate)
```

# Overview

This analysis explores coaching networks and career pathways in NCAA women's basketball, focusing on which schools produce future head coaches and identifying coaching "trees."

Your Task: Work through each section, making decisions and identifying potential story angles from the data.

---

# Data Loading

Load the CSV file and check out the data to see what you're working with.

```{r load-data}
coaching <- read_csv("https://raw.githubusercontent.com/dwillis/wbb/refs/heads/main/ncaa/coaches/coaching_histories_merged.csv")

cat("\nDataset summary:\n")
cat("Total records:", nrow(coaching), "\n")
cat("Unique coaches:", n_distinct(coaching$coach), "\n")
cat("Unique schools:", n_distinct(coaching$college), "\n")
cat("Division I records:", sum(coaching$division == "I", na.rm = TRUE), "\n")
```

---

# Analysis 1: Head Coach Production Pipeline

## DECISION POINT 1: Defining "Future Head Coaches"

**Question for you**: Should we focus only on coaches who became D1 head coaches, or also include those who became head coaches at D2, D3, or other levels? How does this choice affect the story? Does it matter if someone became a D3 head coach vs. stayed in D1 as an assistant?

**Your choice** (modify the code below):

```{r head-coach-pipeline}
# STUDENT DECISION: Modify this filter as needed
# Option A: Only D1 head coaches (current setting)
# Option B: All NCAA head coaches (remove division filter)
# Option C: D1 + D2 only (division %in% c("I", "II"))

d1_head_coaches <- coaching |>
  filter(division %in% c("I", "II"), position_title_standardized == "Head Coach") |>  # MODIFY THIS LINE
  distinct(coach) |>
  pull(coach)

cat("Number of head coaches identified:", length(d1_head_coaches), "\n")
```


## Finding Previous Positions

```{r previous-positions}
# Find their previous positions (before becoming head coach)
future_hc_positions <- coaching |>
  filter(coach %in% d1_head_coaches, position_title_standardized != 'Head Coach') |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    is_head_coach = position_title_standardized == "Head Coach",
    first_hc_year = min(start_year[is_head_coach], na.rm = TRUE),
    before_hc = start_year < first_hc_year | is.na(first_hc_year)
  ) |>
  ungroup() 

cat("Total positions held by future head coaches:", nrow(future_hc_positions), "\n")
```

## DECISION POINT 2: Which Division Matters for Training?

**Question for you**: When looking at which schools produce future head coaches, should we only count their D1 experience, or also their D2/D3/other experience?

```{r production-schools}
# STUDENT DECISION: Modify the division filter
# Current: Only counting D1 positions
# Alternative: Remove division filter to count all positions

hc_production_schools <- future_hc_positions |>
  filter(before_hc) |>  # MODIFY THIS LINE IF NEEDED
  group_by(college_clean, position_title_standardized) |>
  summarize(
    future_hcs = n_distinct(coach),
    .groups = "drop"
  ) |>
  arrange(desc(future_hcs))

cat("\nTop 20 D1 schools producing future head coaches (all positions):\n")
hc_production_schools |>
  group_by(college_clean) |>
  summarize(total_future_hcs = sum(future_hcs), .groups = "drop") |>
  arrange(desc(total_future_hcs)) |>
  head(20) |>
  print(n = 20)
```

**Think about**: Does a stint at a D3 school early in someone's career "count" as part of their development?

## Position Type Breakdown

```{r position-breakdown}
cat("\n\nTop schools by position type:\n")
hc_production_schools |>
  filter(position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")) |>
  group_by(position_title_standardized) |>
  slice_max(future_hcs, n = 10) |>
  arrange(position_title_standardized, desc(future_hcs))
```

### 💡 STORY IDEAS - Section 1

**Based on what you see above, brainstorm 2-3 story angles:**

1. My first thought was focusing on what seems like head pure coaching factories. Mississippi State and Wake Forest lead the list, showing that a small group of stable, well-resourced programs consistently produce future head coaches.
2. After considering this, I'd also be interested to look more into the mid-majors. Schools like VCU, East Carolina, Cincinnati, and Saint Louis appear frequently, suggesting that assistants at these programs gain meaningful experience that could translates into head-coaching jobs. I'm wondering if this is because at mid-majors, they may have to spend more time developing players as they don't always get the star recruits. That's valuable experiences to bring into a new head coaching role, they are used to having to work harder.
3. Finally I'd expand on the assistant coaches dominating the pipeline, it seems like experience could be more valued than high titles. Time spent developing within strong programs seems to matter more than holding an elevated title like Associate Head Coach.

**Questions to consider:**
- Which school dominates the list? Is this surprising?
- Are there mid-major schools punching above their weight?
- Are there regional patterns?

---

# Analysis 2: Career Ladder Analysis

## DECISION POINT 3: Minimum Position Threshold

**Question for you**: Should we require a minimum number of positions before someone counts as having a "career ladder"? (e.g., exclude people who went straight from 1 assistant job to head coach?)

```{r career-ladder}
# STUDENT DECISION: Set minimum number of positions
MIN_POSITIONS <- 2  # CHANGE THIS NUMBER (try 2, 3, or keep at 1)

# Sequence of positions before becoming head coach
career_sequences <- future_hc_positions |>
  filter(before_hc, division == "I") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  summarize(
    positions_held = list(position_title_standardized),
    n_positions = n(),
    final_position = last(position_title_standardized),
    .groups = "drop"
  ) |>
  filter(n_positions >= MIN_POSITIONS)  # APPLIES YOUR THRESHOLD

cat("Coaches meeting threshold:", nrow(career_sequences), "\n")

cat("\nMost common final position before becoming D1 head coach:\n")
career_sequences |>
  count(final_position, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  head(10)

# Average number of positions held before head coach role
cat("\n\nAverage D1 positions held before becoming head coach:",
    round(mean(career_sequences$n_positions, na.rm = TRUE), 1))
```

## Common Career Pathways

```{r pathways}
# STUDENT DECISION: How many positions should we examine?
PATHWAY_LENGTH <- 2  # CHANGE THIS (2, 3, 4, or 5)

cat("\n\nCommon", PATHWAY_LENGTH, "-position pathways to head coach:\n")
career_sequences |>
  filter(n_positions == PATHWAY_LENGTH) |>
  mutate(pathway = map_chr(positions_held, ~paste(.x, collapse = " → "))) |>
  count(pathway, sort = TRUE) |>
  head(15) |>
  print(n = 15)
```

### 💡 STORY IDEAS - Section 2

**Based on the career ladder data, brainstorm 2-3 story angles:**

1. My first angle would be analyzing the long assistant coach grind. The most common pathways are Assistant Coach to Assistant Coach and Assistant Coach to Associate Head Coach, showing that most coaches do not often skip steps but instead spend multiple stints in assistant roles before becoming head coaches.
2. Another angle would be to highlight that head coaching experience is not mandatory. While the Associate Head Coach role appears frequently, many coaches reach head-coaching jobs without ever holding that title, suggesting it is a useful credential but not a necessary one.
3. Finally, staff roles prove to be a good entry point. Paths like Staff to Assistant Coach indicate that non-coaching or support roles can serve as an early foothold in the profession, even if they are rarely the final step before becoming a head coach.

**Questions to consider:**
- Are there "fast track" coaches who skipped steps?
- Does the Associate Head Coach position matter, or do people jump from Assistant to Head?
- How important is the "Staff" level as a starting point?

---

# Analysis 3: Coaching Trees and Mentorship Networks

## DECISION POINT 4: Time Overlap Requirements

**Question for you**: How should we define "worked under"? Should we require they overlapped for at least one full year, or is any overlap sufficient?

```{r coaching-trees}
# STUDENT DECISION: Set minimum overlap threshold
MIN_OVERLAP_YEARS <- 1  # CHANGE THIS (0 = any overlap, 1, 2, 3 for stricter definition)

# Create edges: assistant coaches who worked under head coaches
coaching_relationships <- coaching |>
  filter(division == "I") |>
  select(coach, college_clean, position_title_standardized, start_year, end_year) |>
  # For each school/timeframe, find head coaches
  left_join(
    coaching |>
      filter(division == "I", position_title_standardized == "Head Coach") |>
      select(hc_name = coach, college_clean, 
             hc_start = start_year, hc_end = end_year),
    by = "college_clean",
    relationship = "many-to-many"
  ) |>
  # Keep only overlapping tenures (assistant worked under this head coach)
  filter(
    !is.na(hc_start),
    coach != hc_name,
    position_title_standardized %in% c("Assistant Coach", "Associate Head Coach"),
    # Check for time overlap
    start_year <= coalesce(hc_end, 2025),
    coalesce(end_year, 2025) >= hc_start
  ) |>
  # STUDENT DECISION: Add minimum overlap requirement
  mutate(
    overlap_years = pmin(coalesce(end_year, 2025), coalesce(hc_end, 2025)) - 
                    pmax(start_year, hc_start)
  ) |>
  filter(overlap_years >= MIN_OVERLAP_YEARS) |>  # APPLIES YOUR THRESHOLD
  distinct(coach, hc_name, college_clean)

cat("Coaching relationships identified:", nrow(coaching_relationships), "\n")
```

## Top Mentors

```{r mentors}
# Most prolific mentors (head coaches who produced assistants who became HCs)
prolific_mentors <- coaching_relationships |>
  filter(coach %in% d1_head_coaches) |>
  count(hc_name, sort = TRUE) |>
  rename(mentor = hc_name, proteges_who_became_hc = n)

cat("\nTop 20 mentors (head coaches whose assistants became D1 head coaches):\n")
print(prolific_mentors |> head(20), n = 20)
```

### 💡 STORY IDEAS - Section 3

**Based on the coaching trees data, brainstorm 2-3 story angles:**

1. Ofcourse I have to start with some sort of Brenda Frese incorporation, luckily she is the top mentor. I want to know how coaches like Frese, Kim Barnes Arico, and Kristy Curry have dominate the mentoring game for so long. A story could focus on how their programs consistently function as leadership incubators, not just winning teams. 
2. Still looking into Frese, I think it would be interesting to see how her coaching tree has had a few different destinations in terms of teams and conferences. Profiling her tree could show how philosophies and networks spread far beyond a single Maryland program.
3. Similarly to one of my earlier angles, I can point out the underdogs who had unexpected influence. While some mentors coach at prolific programs, others on the list are less nationally prominent. This angle could explore how effective teachers and program builders can shape the sport just as much as championship-winning head coaches.

**Questions to consider:**
- Who are the most prolific "mentor" coaches?
- Are there surprising names?
- Do certain coaches have "coaching trees" at multiple schools?
- Could you profile a specific coaching tree? Which one?

## DECISION POINT 5: Network Visualization Scope

**Question for you**: How many top mentors should we include in the visualization? Too many = cluttered, too few = missing important connections.

```{r network-visualization, fig.width=12, fig.height=10}
# STUDENT DECISION: Number of top mentors to visualize
N_MENTORS_TO_SHOW <- 15  # CHANGE THIS (try 10, 15, 20, or 25)

# Create network of top mentors (for visualization clarity)
top_mentors <- prolific_mentors |>
  head(N_MENTORS_TO_SHOW) |>
  pull(mentor)

network_edges <- coaching_relationships |>
  filter(
    hc_name %in% top_mentors,
    coach %in% d1_head_coaches
  ) |>
  select(from = hc_name, to = coach, school = college_clean)

# Build network graph
coaching_network <- as_tbl_graph(network_edges, directed = TRUE)

# Plot
ggraph(coaching_network, layout = 'fr') +
  geom_edge_link(aes(alpha = 0.3), arrow = arrow(length = unit(2, 'mm'))) +
  geom_node_point(size = 5, color = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  labs(
    title = "Coaching Trees: Top Mentors and Their Protégés",
    subtitle = paste("Top", N_MENTORS_TO_SHOW, "D1 head coaches who mentored future D1 head coaches")
  )
```

---

# Analysis 4: School-to-School Coaching Pipelines

```{r school-pipelines}
# Track coach movement between schools
coach_movements <- coaching |>
  filter(division == "I") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_school = lead(college_clean),
    next_position = lead(position_title_standardized),
    next_start = lead(start_year)
  ) |>
  filter(!is.na(next_school), college_clean != next_school) |>
  ungroup()

# Most common school-to-school transitions
cat("\nTop 30 school-to-school coaching pipelines:\n")
coach_movements |>
  count(college_clean, next_school, sort = TRUE) |>
  rename(from_school = college_clean, to_school = next_school, coaches = n) |>
  head(30) |>
  print(n = 30)
```

## Schools that Export to Head Coaching

```{r exports}
# Schools that "export" coaches to D1 head coaching jobs
coach_exports <- coach_movements |>
  filter(next_position == "Head Coach") |>
  count(college_clean, sort = TRUE) |>
  rename(school = college_clean, coaches_to_hc = n)

cat("\n\nTop 20 schools that export coaches to D1 head coaching positions:\n")
print(coach_exports |> head(20), n = 20)
```

### 💡 STORY IDEAS - Section 4

**Based on school-to-school movement patterns, brainstorm 2-3 story angles:**

1. The first story that jumps out are regional and conference based pipelines. Many of the strongest school-to-school connections are regional like Mississippi State to Texas, Memphis to Cinci, and Marshall to Tennessee. This suggests that geographic proximity and conference familiarity play a major role in coaching mobility.
2. The next thing that jumps out to me are the unexpected long-distance pipelines. Pipelines like Arizona to SMU help to highlight the importance of personal networks and coaching relationships over location alone.
3. Another angle could be power programs serving as feeder schools to mid-majors. Several high-profile programs appear frequently as exporters of coaches, while the receiving schools are often mid-major or stepping-stone programs. This points to a common career pattern where assistants leave elite staffs to gain head-coaching experience elsewhere.

**Questions to consider:**
- Are there strong regional pipelines (e.g., schools from the same conference feeding each other)?
- Do assistant coaches at top programs tend to move to head coaching at mid-majors first?
- Are there unusual cross-country connections?
- Which schools are "feeders" vs. "destinations"?

---

# Analysis 5: Conference-Level Analysis

## DECISION POINT 6: Conference Focus

**Question for you**: Should we analyze all conferences equally, or focus on major/Power 5 conferences? Should we combine some smaller conferences?

```{r conference-analysis}
# STUDENT DECISION: Define which conferences to highlight
# Option A: Analyze all conferences (current)
# Option B: Create a "Power 5" vs "Mid-Major" vs "Small" grouping
# Option C: Focus only on specific conferences

# Conferences producing future head coaches
conference_production <- future_hc_positions |>
  filter(before_hc, division == "I") |>
   mutate(
    conference_tier = case_when(
      conference %in% c("ACC", "Big Ten", "Big 12", "Pac-12", "SEC") ~ "Power 5",
      conference %in% c(
        "AAC", "A-10", "Atlantic 10", "Big East", "C-USA", "Conference USA",
        "MAC", "Mountain West", "WCC", "Sun Belt", "WAC"
      ) ~ "Mid-Major",
      TRUE ~ "Small"
    )
  ) |>
  count(conference_tier, position_title_standardized) |>
  arrange(desc(n))

cat("\nTop conferences producing future D1 head coaches:\n")
conference_production |>
  group_by(conference_tier) |>
  summarize(total = sum(n), .groups = "drop") |>
  arrange(desc(total)) |>
  head(15) |>
  print(n = 15)
```

## Conference Loyalty

```{r loyalty}
# Conference loyalty: coaches who stay within conference
conference_loyalty <- coach_movements |>
  left_join(
    coaching |> 
      filter(division == "I") |>
      select(coach, college_clean, conference),
    by = c("next_school" = "college_clean", "coach" = "coach"),
    relationship = "many-to-many"
  ) |>
  filter(!is.na(conference.x), !is.na(conference.y)) |>
  mutate(same_conference = conference.x == conference.y) |>
  group_by(conference.x) |>
  summarize(
    total_moves = n(),
    same_conference_moves = sum(same_conference),
    loyalty_rate = same_conference_moves / total_moves * 100,
    .groups = "drop"
  ) |>
  arrange(desc(loyalty_rate))

cat("\n\nConference retention rates (coaches staying within conference):\n")
print(conference_loyalty |> head(15), n = 15)
```

### 💡 STORY IDEAS - Section 5

**Based on conference patterns, brainstorm 2-3 story angles:**

1. Small conferences seem to serve as coaching incubators as they and mid-majors collectively produce more future head coaches than Power 5 conferences. Lower-resource environments may offer assistants broader responsibilities and clearer paths to advancement which is similar to what I alluded to earlier.
2. There is low conference loyalty across the board. Even in the most “loyal” leagues, fewer than 40% of coaching moves stay within the same conference, indicating that career advancement typically requires geographic and conference mobility rather than staying put.
3. Lastly, Different conferences play different roles. Conferences like the SEC and Pac-12 generate many coaching moves but retain relatively few internally, positioning them as exporters of coaching talent, while smaller leagues such as the SWAC and Big Sky show higher retention and may function more as internal development systems. 

**Questions to consider:**
- Do certain conferences "develop" more coaches than others?
- Is there high loyalty within conferences, or do coaches move around?
- Are there conferences that import talent vs. export it?

---

# Analysis 6: Time to Head Coach Role

## DECISION POINT 7: Career Start Definition

**Question for you**: When calculating "time to head coach," should we count ALL positions (including high school, community college) or only college-level positions?

```{r time-to-hc}
# STUDENT DECISION: Filter on category
# Current: All positions included
# Alternative: Add filter like "category == 'College'" to exclude high school, etc.

time_to_hc <- coaching |>
  filter(coach %in% d1_head_coaches) |>
   filter(category == "College") |>  # UNCOMMENT THIS LINE TO EXCLUDE NON-COLLEGE
  arrange(coach, start_year) |>
  group_by(coach) |>
  summarize(
    first_position_year = min(start_year, na.rm = TRUE),
    first_hc_year = min(start_year[position_title_standardized == "Head Coach" & 
                                     division == "I"], na.rm = TRUE),
    years_to_hc = first_hc_year - first_position_year,
    .groups = "drop"
  ) |>
  filter(is.finite(years_to_hc), years_to_hc >= 0)

cat("\nTime to D1 head coach position:\n")
cat("Mean:", round(mean(time_to_hc$years_to_hc, na.rm = TRUE), 1), "years\n")
cat("Median:", median(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")
cat("Min:", min(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")
cat("Max:", max(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")

# Distribution
ggplot(time_to_hc, aes(x = years_to_hc)) +
  geom_histogram(binwidth = 2, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = median(time_to_hc$years_to_hc, na.rm = TRUE), 
             linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Years to First D1 Head Coaching Position",
    subtitle = "From first coaching role to D1 head coach",
    x = "Years",
    y = "Count of coaches"
  ) +
  theme_minimal()
```

### 💡 STORY IDEAS - Section 6

**Based on career timeline data, brainstorm 2-3 story angles:**

1. I want to know more about the 10-12 year grin in takes to become a head coach. This challenges the idea of rapid ascension in college sports. Most coaches spend years in assistant and support roles before ever getting a chance to lead. The data suggests success is more about persistence than speed.
2. Some coaches reach D1 head coach roles in just a few years, while others take two decades or more. These divergent timelines point to fundamentally different career paths. Are there elite playing resumes, strong networks, or timely promotions versus long-term apprenticeships? There is no single formula for reaching the top.
3. A significant number of coaches don’t get their first D1 head coaching job until 20 or more years into their careers. For many, the opportunity comes late and may be their only shot. This raises questions about access, timing, and how opportunity is distributed in college coaching.

**Questions to consider:**
- What's the typical timeline to becoming a D1 head coach?
- Who are the outliers (very fast or very slow)?

---

# Analysis 7: Elite Program "Finishing Schools"

## DECISION POINT 8: Defining "Elite Programs"

**Question for you**: Which schools should count as "elite" or "blue blood" programs? 
This is subjective - modify the list based on your knowledge or research.

```{r finishing-schools}
# STUDENT DECISION: Modify this list of top programs
# Consider: recent success, historical success, resources, national profile

top_programs <- c("UConn", "Tennessee", "Stanford", "Notre Dame", "Baylor",
  "South Carolina", "LSU", "Texas", "UCLA",
  "Louisville", "Maryland", "Ohio St.", "Duke")
# ADD OR REMOVE SCHOOLS AS YOU SEE FIT

cat("Analyzing", length(top_programs), "programs\n")

# Assistant coaches from top programs who became head coaches
finishing_school_analysis <- future_hc_positions |>
  filter(
    before_hc,
    division == "I",
    college_clean %in% top_programs,
    position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")
  ) |>
  group_by(college_clean) |>
  summarize(
    future_head_coaches = n_distinct(coach),
    .groups = "drop"
  ) |>
  arrange(desc(future_head_coaches))

cat("\n'Finishing schools' - top programs producing future head coaches:\n")
print(finishing_school_analysis, n = Inf)

# Average years spent at top program before becoming HC
years_at_top_programs <- future_hc_positions |>
  filter(
    before_hc,
    division == "I",
    college_clean %in% top_programs,
    position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")
  ) |>
  mutate(years_at_school = coalesce(end_year, 2025) - start_year) |>
  group_by(college_clean) |>
  summarize(
    avg_years = mean(years_at_school, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_years))

cat("\n\nAverage years spent at top programs (for future HCs):\n")
print(years_at_top_programs, n = Inf)
```

### 💡 STORY IDEAS - Section 7

**Based on elite program patterns, brainstorm 2-3 story angles:**

1. Elite programs seem to serve as true finishing schools. Programs like UConn, Notre Dame, and Baylor show long average assistant tenures before coaches become head coaches, suggesting these schools function as finishing schools where extended apprenticeships prepare assistants for top jobs.
2.This data proves development is more important than dominance if you want to be the next big thing. Programs that give assistants time, responsibility, and mentorship appear to prepare coaches for head roles more effectively than those that simply accumulate talent and wins.
3. Some blue-blood programs appear to function as rapid credentialing platforms, where the program’s name and visibility help assistants secure head coaching jobs after brief stints. Duke (8 future HCs, 1.85 years), LSU (3, 1.67), and Texas (9, 2.3) illustrate this effect: assistants leverage the program’s reputation to move quickly into top roles.

**Questions to consider:**
- Do the biggest brand names produce the most head coaches?
- Are there "hidden gems" - programs that punch above their weight?
- Does time spent at elite programs matter (short stint vs. long apprenticeship)?
- Are there programs that hoard talent vs. those that develop and release it?

---

# Analysis 8: "Stepping Stone" Schools

```{r stepping-stones}
# Head coaches who left for another D1 head coaching job
stepping_stone_moves <- coaching |>
  filter(division == "I", position_title_standardized == "Head Coach") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_school = lead(college_clean),
    next_position = lead(position_title_standardized),
    moved_to_hc = next_position == "Head Coach" & !is.na(next_school)
  ) |>
  filter(moved_to_hc) |>
  ungroup()

cat("\nSchools most commonly serving as stepping stones (HC → HC moves):\n")
stepping_stone_moves |>
  count(college_clean, sort = TRUE) |>
  rename(school = college_clean, hc_departures = n) |>
  head(20) |>
  print(n = 20)

# Most common destination schools
cat("\n\nMost common destination schools for head coaches:\n")
stepping_stone_moves |>
  count(next_school, sort = TRUE) |>
  rename(destination = next_school, hc_arrivals = n) |>
  head(20) |>
  print(n = 20)
```

### 💡 STORY IDEAS - Section 8

**Based on stepping stone patterns, brainstorm 2-3 story angles:**

1.Schools like South Dakota, Albany, Charlotte, and Missouri State appear repeatedly as places where successful head coaches leave for other D1 jobs, marking them as true stepping-stone programs rather than long-term destinations.
2. Some of the most common arrival schools (Wisconsin, Texas Tech, Auburn) tend to have more resources and visibility, reinforcing the idea that head-coaching careers often move upward through a recognizable institutional ladder.

**Questions to consider:**
- Which schools can't keep their successful head coaches?
- What makes a school a "destination" vs. a "stepping stone"?
- Are there schools that serve as consistent "first head coaching jobs"?
- Can you identify a "ladder" of schools (small to medium to major)?

---

# Analysis 9: Position Mobility Analysis

```{r position-mobility}
# Years spent in each position before moving up
position_duration <- coaching |>
  filter(division == "I") |>
  mutate(years_in_position = coalesce(end_year, 2025) - start_year) |>
  group_by(position_title_standardized) |>
  summarize(
    n_positions = n(),
    avg_duration = mean(years_in_position, na.rm = TRUE),
    median_duration = median(years_in_position, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(n_positions))

cat("\nAverage years in each position type:\n")
print(position_duration, n = Inf)

# Promotion rates by position
promotions <- coaching |>
  filter(division == "I") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_position = lead(position_title_standardized),
    promoted = case_when(
      position_title_standardized == "Staff" & 
        next_position %in% c("Assistant Coach", "Associate Head Coach", "Head Coach") ~ TRUE,
      position_title_standardized == "Assistant Coach" & 
        next_position %in% c("Associate Head Coach", "Head Coach") ~ TRUE,
      position_title_standardized == "Associate Head Coach" & 
        next_position == "Head Coach" ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  filter(!is.na(next_position)) |>
  group_by(position_title_standardized) |>
  summarize(
    total_moves = n(),
    promotions = sum(promoted),
    promotion_rate = promotions / total_moves * 100,
    .groups = "drop"
  ) |>
  arrange(desc(promotion_rate))

cat("\n\nPromotion rates by position:\n")
print(promotions, n = Inf)
```

### 💡 STORY IDEAS - Section 9

**Based on mobility and promotion patterns, brainstorm 2-3 story angles:**

1. Entry-level roles are crucial launching pads. Spending 2 years as a staffer can lead to an Assistant Coach or higher position, making these roles a high-leverage entry point for aspiring coaches.
2. As coaches climb the ladder, promotions become harder and slower. There’s a “sweet spot” of time in each role—long enough to gain experience, but not so long that upward mobility stagnates.
3. Some roles represent the pinnacle or a side track. Once in a Head Coach or specialized position, upward mobility stalls, highlighting the hierarchy and structural ceilings within coaching careers.

**Questions to consider:**
- Which positions are "dead ends" vs. "springboards"?
- Do people spend more time in certain positions?
- Is there a "right" amount of time to spend before moving up?

---

# FINAL REFLECTION: Your Best Story Ideas

Review all your story ideas from each section. Which 3 would you pursue as full stories?

**Story #1:**
Elite programs as finishing schools vs. rapid credentialing platforms (combining ideas 1 and 3 from section 7) This looks at the idea that some schools cultivate coaches through long apprenticeships (UCONN, Notre Dame, Baylor) while others leverage brand and prestige for faster promotion (Duke, Texas, LSU)

**Why this story:**
Understanding these different pathways reveals how coaching talent is cultivated and how brand reputation can substitute for time in developing skills. This distinction helps explain why some assistants rise quickly while others benefit from years of preparation.

---

**Story #2:**
The importance of stepping-stone and mid-major programs in career development (this combines setions 1 and 8) Mid-major and smaller programs act as critical training grounds where assistants gain broad responsibility before moving to head-coaching roles elsewhere.

**Why this story:**
Mid-major and smaller programs often provide broader responsibilities and hands-on experience that assistants might not get at elite schools. Highlighting their role challenges the notion that only top-tier schools produce top-tier coaches.

---

**Story #3:**
Position mobility and promotion patterns (combining stories 1 and 2 from section 9) The data on Staff, Assistant, and Associate Head Coach roles provides a clear story about how early-career positions serve as launchpads, the “sweet spot” for experience, and also where mobility can stall.

**Why this story:**
Examining promotion patterns shows where upward mobility is most likely and where career ceilings appear, giving insight into how experience and timing shape coaching trajectories. This helps aspiring coaches understand the “sweet spot” for gaining skills before moving up.

---

# Additional Analysis Ideas to Explore

Now that you've worked through these analyses, what other questions could this data answer?

**Your ideas:**

1. A diversity/representation analysis:
For this I would analyze demographic patterns in career progression, for example gender, race, or former playing experience, and whether certain groups have faster or slower pathways to head-coaching roles. This could reveal structural advantages or barriers.
2. Impact of program success on career trajectory:
To achieve this I could examine whether assistants at highly successful programs (conference champions, frequent NCAA tournament appearances) advance faster than those at struggling programs, controlling for tenure. This could quantify how much winning versus development matters in climbing the ladder.
3. Hidden Gems:
For this I would need to identify mid-major programs that consistently produce future head coaches relative to their resources and profile. I think this could create a nice bubble chart showing number of future HCs versus program prestige. It could highlight programs punching above their weight.