## NCAA Football Attendance 

Let's load the tidyverse.

```{r}
library(tidyverse)
```

And read in the attendance data:

```{r}
attendance <- read_csv("https://dwillis.github.io/sports-data-files/attendance.csv")
```

Using the tidyverse's `pivot_longer` function, turn the currently wide data into longer data and save it as a new dataframe called `attendance_longer`:

```{r}
attendance_longer <- attendance |> 
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "Attendance")
```

Let's try to figure out whether football attendance has returned to "normal" by 2024, defining "normal" as 2019 figures. For that we'll need to isolate those two years using `filter()` and get rid of any NAs:

```{r}
attendance_two_years <- attendance_longer |> filter(Year == 2019 | Year == 2024) |> 
  na.omit()
```

Now, in order to calculate a percentage change between those two years, we'll need to go back to wider data and again ditch teams where we don't have both years:

```{r}
attendance_wide <- attendance_two_years |> pivot_wider(names_from = Year, values_from = Attendance) |> na.omit()
```

Finally, we'll add a percentage change column to that dataframe (remember, it's (New-Old)/Old) and arrange it to see the results.

```{r}
attendance_wide <- attendance_wide |> mutate(percent_change = (`2024` - `2019`)/`2019` * 100) |> 
  arrange(percent_change)
```

Some of these clearly are influenced by performance on the field. Below, describe the most interesting results (positive or negative) and what you think might be going on with them. Feel free to do some googling and describe what other data would be helpful to clarify what the best stories here might be. What surprises you about these results?

Similarly to last time, SMU emerges as one of the most interesting cases. In 2019 the Mustangs already drew healthy crowds, but after pivoting the data to compare only 2019 and 2024, their attendance shows one of the steepest positive trajectories in the entire dataset with a 61.2% percent increase. This increase was by far the largest among Power Four programs, highlighting that their rebound was not just a return to pre-pandemic levels but a dramatic expansion of fan interest. I think several factors could be to blame including a strong 2023 campaign that kept the team in national conversations, the high-profile announcement and execution of their move to the ACC beginning in 2024, and aggressive marketing and facility upgrades in Dallas that encouraged larger, more consistent crowds. SMU also had a change in management bringing in a new Athletic Director and Executive Deputy Athletic Director so I'm interested to see what their 2025 attendance demonstrates later. Together these elements show how a combination of athletic success, conference changes, and institutional investment can translate directly into growth in game-day attendance.



