---
title: "Women's Soccer Data"
---

We'll use this notebook to explore some NCAA women's soccer data and [our questions](https://docs.google.com/spreadsheets/d/1G1AjH2iS6kj3kKHCRi0XVy5Lcm-rahJkeTAToVh9J1g/edit?gid=0#gid=0). First, let's load the tidyverse library:

```{r}
library(tidyverse)
```

We'll want to start with 2025 match data, which is located in [this GitHub repository](https://github.com/dwillis/NCAAWomensSoccerData). We can load the match data directly from GitHub using `read_csv`:


```{r}
matches_2025 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2025.csv")
```

Let's take a look at this data to see what it contains, using `head` and `glimpse`:

```{r}
head(matches_2025)
glimpse(matches_2025)
```

Now, onto the questions using aggregates, which means `group_by` and `summarize`:

1. How many shots in total has Maryland attempted and how many have been on goal? Calculate a percentage of shots on goal.

Maryland attempted 83 shots and 45 of them were on goal. 54.2% was the percentages of shots on goal.

```{r}
matches_2025 |>
  group_by(team) |>
  summarise (
    shots_att = sum(sh_att),
    sog = sum(so_g),
    sht_pct = (sog/shots_att)*100
  )
```


2. What about the same shot stats from last season's first four games? You'll need to load the matches from 2024 - just change the season in the URL - and filter by date to find the first four games.

In the first four games, Maryland attempted 53 shots with 27 of them being on goal. Their shot on goal percentage through the first four games was 50.9%.

```{r}
matches_2024 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv")

matches_2024 |>
    group_by(team="Maryland Terrapins, Terps") |>
     arrange(date) |>
  head(n=4) |>
 summarise (
    shots_att = sum(sh_att),
    sog = sum(so_g),
    sht_pct = (sog/shots_att)*100 
    )
```  


3. Maryland started 2022 with 4 draws. Have any other teams had as many this season?
Yes, several teams have four or more draws this season, the highest number of draws came from Drake and Navy at seven.

```{r}
matches_2025 |>
  group_by(team, outcome = "draw") |>
  summarise (
    total = n ()
  ) |>
  arrange(desc(total))
```


4. How many total corner kicks has Maryland had so far this season and how many have its opponents have? Calculate the difference between the two totals for every team using group_by, summarize and mutate, then filter to find Maryland's totals.

Maryland has had 28 corner kicks this season while their opponents have only had 22. The difference is 6 kicks.

```{r}
matches_2025 |>
  group_by(team) |>
  summarise (
    corner_kicks = sum(corners),
    opp_corner_kicks = sum(defensive_corners)
  ) |>
  mutate(corner_diff = corner_kicks - opp_corner_kicks)  |>
  filter(team=="Maryland Terrapins, Terps")

```

5. What's the most interesting thing you've seen from these questions?
I find it really interesting that Maryland had only 53 shots in the first four games of 2024 but already has 83 shots in 2025 with just two more games played. That’s a jump of 20 shots in only two extra games, showing a big increase in their offensive activity. It’s impressive how they’re creating even more scoring opportunities this season while maintaining solid accuracy. Their shot on goal percentage has gone up by almost 4%  while taking more shots. It makes me wonder if they are taking smarter shots or the talent is the causation. 

```{r}
swac_teams <- c("Alabama A&M Bulldogs", "Alabama St. Lady Hornets", "Alcorn Lady Braves", "Ark.-Pine Bluff Lady Lions", "Grambling Lady Tigers", "Jackson St. Lady Tigers", "Mississippi Val. Devilettes", "Prairie View Lady Panthers", "Southern U. Lady Jaguars", "Texas Southern Lady Tigers")

swac_matches <- matches_2025 |> 
  filter(team %in% swac_teams)

matches_2025 |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

swac_matches |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

big10 <- c("Nebraska Huskers", "Iowa Hawkeyes", "Minnesota Golden Gophers", "Illinois Fighting Illini", "Northwestern Wildcats", "Wisconsin Badgers", "Indiana Hoosiers", "Purdue Boilermakers", "Ohio St. Buckeyes", "Michigan Wolverines", "Michigan St. Spartans", "Penn St. Nittany Lions", "Rutgers Scarlet Knights", "Maryland Terrapins, Terps", "Southern California Trojans", "UCLA Bruins", "Washington Huskies", "Oregon Ducks")

big10_matches <- matches_2025 |> 
  filter(team %in% big10)

big10_matches |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

big10_matches |> 
  filter(team == "Maryland Terrapins, Terps") |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )
```

Significance Testing
```{r}
matches_2024 <- matches_2024 |> 
  mutate(total_goals =  goals + defensive_goals)

matches_2025 <- matches_2025 |> 
  mutate(total_goals =  goals + defensive_goals)

t.test(matches_2024$total_goals, matches_2025$total_goals)
```
Results:
Based on the t-test output, the mean total goals increased from 2.75 in 2024 to 2.90 in 2025, indicating scoring is slightly up this year. The p-value, 0.00047, is well below 0.05, so this difference is statistically significant. The confidence interval, -0.236 to -0.067, does not include 0, further confirming a significant increase. This change could be due to factors like stronger offensive play, weaker defenses, or the uneven matchups. The season is just starting so we are not well into conference play. As teams continue to play tougher opponents, I anticipate their average goals to decrease slightly. 

Significance Testing for UMD games.
```{r}
maryland_2024 <- matches_2024 |> 
  filter(team == "Maryland Terrapins, Terps") |> 
  mutate(total_goals =  goals + defensive_goals)

maryland_2025 <- matches_2025 |> 
  filter(team == "Maryland Terrapins, Terps") |> 
  mutate(total_goals =  goals + defensive_goals)

t.test(maryland_2024$total_goals, maryland_2025$total_goals)

```

Results:
For Maryland games, the average total goals increased slightly from 2.68 in 2024 to 2.88 in 2025, but the p-value, 0.766, is much higher than 0.05, so this difference is not statistically significant. The confidence interval, -1.54 to 1.16, includes 0, further indicating no significant change. This suggests that, so far, Maryland’s scoring is roughly the same as last year. Since the 2025 season is not yet halfway complete, the results could change as more games are played. Additionally, since the Terps generally perform better against non-conference opponents and have only played one conference game so far, their scoring average may not yet reflect typical performance across the full season.

