---
title: "background"
format: html
---

Question: Does the number of turnovers by Maryland’s team influence the number of goals allowed in a lacrosse game?

1. Does that data already exist? Can you get it?
  - Yes, this data does exist
  - These statistics are tracked in databases like stats.ncaa.org
  - I am copying the “Game by Game Stats” table into a google sheet document
  - Then cleaning the data: cleaning opponent names, modifying the “Result” column to “W” or “L,” Adding a “Location” column with “home,” “away,” or “neutral,” and finally adding a “Goals_Allowed” column
  - I can then save my document as a csv file to work with

2. Do you need assistance in getting it?
  - Yes, I am not sure how to track possession time which I think will add context to my analysis

3. What time frame do you want/need data from?
  - I am looking to compile data from the 2021 season - present
  - I want enough data where I can rule out one-season anomalies and demonstrate a pattern

4.Will you have to improve or augment the data in order to answer your question? If so, how?
  - Yes I want to add the following contextual variables
    - home/away
    - Measure the correlation between goals allowed per turnover

5. Are there other examples of this kind of work (specific or general)? If so, provide them.
  - Similar studies have been done in other sports
    - Basketball – turnover margin vs. points allowed
    - Football - turnover differential vs. win probability
```{r}
library(tidyverse)
```

```{r}
season25 <- read_csv("/Users/lornizzle/Downloads/2025 lax data - Sheet1.csv")
season24 <- read_csv("/Users/lornizzle/Downloads/2024 lax data - Sheet1.csv")
season23 <- read_csv("/Users/lornizzle/Downloads/2023 lax data - Sheet1.csv")
season22 <- read_csv("/Users/lornizzle/Downloads/2022 lax data - Sheet1.csv")
season21 <- read_csv("/Users/lornizzle/Downloads/2021 lax data - Sheet1.csv")
```
```{r}
maryland_lax_data <- bind_rows(season21, season22, season23, season24, season25)
```

Data Exploration

1. Describe the data
```{r}
nrow(maryland_lax_data)
maryland_lax_data <- maryland_lax_data %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))
range(maryland_lax_data$Date, na.rm = TRUE)
```
I combined all observed seasons into one data-frame and it has 85 rows ranging from Feb. 20, 2021 through May 26, 2025.

2. How many games per opponent? - using this as my key column to check the balance of my data

```{r}
opponent_summary <- maryland_lax_data |> 
  group_by(Opponent) |> 
  summarize(Games_Played = n()) |> 
  arrange(desc(Games_Played))
```

This makes me think I need to observe more seasons as this does not present a lot of games played per opponent. I think my data is not as balanced or complete.

```{r}
ggplot(maryland_lax_data, aes(x = Date, y = TO)) +
  geom_line(color = "red") +
  scale_y_continuous(limits = c(0, 23)) +
  labs(title = "Turnovers Over Time (2021–2025)", x = "Date", y = "Turnovers") +
  theme_minimal()
```

One challenge is that some seasons include additional columns like "FO_Won" and "FOsTaken", which are missing in earlier years, making it harder to compare all seasons consistently. Another issue is the uneven number of games against each opponent, which could bias any relationship between turnovers and goals allowed. Finally, inconsistent or improperly formatted dates initially made it difficult to accurately plot trends over time, adding extra steps before analysis.Small sample sizes for certain opponents further complicate the analysis, as it is difficult to draw meaningful conclusions from only one or two games. These combined factors make it more challenging to definitively answer whether turnovers directly influence goals allowed, and any conclusions must be interpreted with caution.
