---
title: "presentation"
output: html_document
---

Many people have the common belief that in men's lacrosse, turnovers allude to more defensive pressure which would result in more goals allowed. I had the same belief so I decided to take a look at the Terps' performance between 2021 and 2025 to evaluate the following:

Does the number of turnovers in a game influence how many goals Maryland allows?

There is ample data on turnovers but I have rarely, if ever, seen the statistic's observed impact on a teams' goal allowance.

```{r}
library(tidyverse)
```
```{r}
mdseason25 <- read_csv("2025mdlaxdata.csv")
mdseason24 <- read_csv("2024mdlaxdata.csv")
mdseason23 <- read_csv("2023mdlaxdata.csv")
mdseason22 <- read_csv("2022mdlaxdata.csv")
mdseason21 <- read_csv("2021mdlaxdata.csv")

maryland_lax_data <- bind_rows(mdseason21, mdseason22, mdseason23, mdseason24, mdseason25) |> 
  mutate(Date = mdy(Date),
         TO = as.numeric(TO),
         Goals_Allowed = as.numeric(Goals_Allowed),
         Location = as.factor(Location),
         Result = as.factor(Result))
```
I used data from stats.ncaa.org to compile Maryland Men's lacrosse data, then combined the seasons to have all data in one frame. While I initially aimed to include time of possession, this data was not reported in the same game-by-game display so I focused on turnovers and goals allowed. 
```{r}
ggplot(maryland_lax_data, aes(x = TO, y = Goals_Allowed)) +
geom_point(color = "red", size = 3, alpha = 0.7) +
geom_smooth(method = "lm", se = FALSE, color = "black") +
labs(
title = "Relationship Between Turnovers and Goals Allowed (Maryland 2021–2025)",
x = "Turnovers",
y = "Goals Allowed"
) +
theme_minimal()
```
This scatterplot shows a slightly negative relationship between turnovers and games allowed meaning as turnovers increase, the goals allowed actually tend to decrease.
```{r}
cor(maryland_lax_data$TO, maryland_lax_data$Goals_Allowed)

model <- lm(Goals_Allowed ~ TO, data = maryland_lax_data)
summary(model)
```
My regression analysis proved that the relationship is statistically significant so turnovers alone do not strongly predict how many goals Maryland allows.

This led me to ask whether turnovers matter more in determining outcomes rather than defensive goals allowed.

```{r}
ggplot(maryland_lax_data, aes(x = Result, y = TO, fill = Result)) +
geom_boxplot() +
labs(title = "Turnovers by Game Result (Win vs. Loss)",
x = "Result", y = "Turnovers") +
theme_minimal()
```
As expected, more turnovers occur in losses than in games resulting in a Terps win. Although turnovers do not help to predict defensive goals allowed, they do matter and the effect shows up in game outcomes.

To test whether this pattern is unique to Maryland, I compared them to a similar program and longstanding lacrosse rival, Johns Hopkins. 

```{r}
hopseason25 <- read_csv("2025hoplaxdata.csv")
hopseason24 <- read_csv("2024hoplaxdata.csv")
hopseason23 <- read_csv("2023hoplaxdata.csv")
hopseason22 <- read_csv("2022hoplaxdata.csv")
hopseason21 <- read_csv("2021hoplaxdata.csv")

hopkins_lax_data <- bind_rows(hopseason21, hopseason22, hopseason23, hopseason24, hopseason25) |> 
  mutate(Date = mdy(Date),
         TO = as.numeric(TO),
         Goals_Allowed = as.numeric(Goals_Allowed),
         Location = as.factor(Location),
         Result = as.factor(Result))
```
To begin my comparison, I did the same thing for Hopkins as I did to set up Maryland. Now I'm taking it a step further and aggregating both teams and their respective seasons into one dataframe. 
```{r}
all_lax_data <- bind_rows(
  maryland_lax_data  |>  mutate(Team = "Maryland"),
  hopkins_lax_data |>  mutate(Team = "Johns Hopkins"))
```
```{r}
all_lax_data |> 
  group_by(Team)  |> 
  summarise(
    Avg_Turnovers = mean(TO, na.rm = TRUE),
    Avg_Goals_Allowed = mean(Goals_Allowed, na.rm = TRUE)
  ) |> 
  pivot_longer(
    cols = c(Avg_Turnovers, Avg_Goals_Allowed),
    names_to = "Metric",
    values_to = "Value"
  ) |> 
ggplot(aes(x = Team, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  labs(title = "Average Turnovers vs. Goals Allowed (Maryland vs. Hopkins)",
       y = "Average per Game", x = "Team") +
  scale_fill_manual(values = c("Avg_Turnovers" = "blue", "Avg_Goals_Allowed" = "orange")) +
  theme_minimal()
```
Hopkins averages both more turnovers and more goals allowed. While this suggests Maryland has a slight edge when it comes to defensive efficiency, this alone does not prove a relationship between Hopkins' turnovers and goals allowed.
```{r}
hopkins_model <- lm(Goals_Allowed ~ TO, data = hopkins_lax_data)
summary(hopkins_model)
```
Hopkins also shows no meaningful relationship between turnovers and goals allowed confirming, across both programs, turnovers alone are not a reliable predictor of defensive performance. 

So the answer to my question: Does the number of turnovers influence goals allowed? is no, not in a consistent or meaningful way.

Turnovers affect winning, but do not directly determine defensive breakdowns. Next time I would I want to find a way to evaluate opponent strength and goalie save percentages alongside turnovers. 